---
title: "Grace Final Draft 3: Investigating Police Citations in Houston"
subtitle: "Draft 3 - dplyr"
author: "Grace Wang, Virginia Baskin, Siddhi Narayan"
date: "2023-02-26"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load in data, echo=FALSE}

library(ggplot2)
library(ggExtra)
library(ggthemes)
library(readr)
library(dplyr)
mini_htx <- read_csv("Datasets/mini_htx.csv")
mini_htx <- mini_htx %>% 
       rename("Beats" = "beat")

beats_data <- read_csv("Datasets/COH_POLICE_BEATS.csv")
htx_stats <- read_csv("Datasets/DECENNIALPL2020.P2-2023-03-02T224324.csv")

```

## Introduction 

# What is the race/sex breakdown of each type of citation?

Looking at our data, our team noticed that there were 5 main types of violations: speeding, invalid license, failure to establish financial responsibility, failure to wear a seat belt, and running a stop sign/red light. As such, we employed the dplyr library to view the racial breakdown of each type of citation.

Speeding Citations by Race:
```{r q1, echo=FALSE}

# SPEEDING Citations
speeding_race <- mini_htx %>% filter(grepl("SPEEDING", violation)) %>%
         group_by(subject_race) %>%
         summarise(n_individuals = n())
speeding_race
```

Invalid License Citations by Race:
```{r q1, echo=FALSE}
# Invalid License Citations
license_race <- mini_htx %>% filter(grepl("LICENSE", violation)) %>%
         group_by(subject_race) %>%
         summarise(n_individuals = n())
license_race
```

Failure to Establish Financial Responsibility Citations by Race:
```{r q1, echo=FALSE}
# Failure to establish Financial Responsibility Citations
financial_race <- mini_htx %>% filter(grepl("FINANCIAL RESPONSIBILITY", violation)) %>%
         group_by(subject_race) %>%
         summarise(n_individuals = n())
financial_race

```

Seat Belt Citations by Race:
```{r q1, echo=FALSE}
# seat belt citations
seat_belt_race <- mini_htx %>% filter(grepl("SEAT BELT", violation)) %>%
         group_by(subject_race) %>%
         summarise(n_individuals = n())
seat_belt_race

```

Running a Red Light/Stop Sign Citations by Race:
```{r q1, echo=FALSE}
# Running a stop light/red light
stop_light_race <- mini_htx %>% 
  filter(grepl("RED LIGHT", violation)|grepl("STOP SIGN", violation)) %>%
  group_by(subject_race) %>%
  summarise(n_individuals = n())
stop_light_race

```

# What is the number of citations per square mile for each beat?
Next, our team aimed to understand the geographic distribution of citations by calculating the number of citations per square mile within each beat. 

```{r, echo=FALSE}

# join the two dataframes on beats

citations_in_beat <- mini_htx %>% inner_join(beats_data, by="Beats") %>%
  group_by(Beats) %>%
  summarise(n_individuals = n())

citations_in_beat <- as.data.frame(citations_in_beat)

cit_beat_area <- citations_in_beat %>% 
  inner_join(beats_data[,c("Beats", "Area_sq_mi")], by="Beats")

cit_beat_area$Citations_per_sqm <- cit_beat_area$n_individuals / cit_beat_area$Area_sq_mi

cit_beat_area[,c("Beats","Citations_per_sqm")]
```